---
title: "janitor: an overview"
subtitle: "Useful Tools for Cleaning and Exploring Dirty Data"
author: "presented by Gülce Sena Tuncer & Nassim Zoueini"
output: 
  html_document:
    toc: TRUE
    df_print: paged
    number_sections: FALSE
    highlight: tango
    theme: lumen
    toc_depth: 3
    toc_float: true
    css: custom.css 
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=2)

colorise <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```

Welcome to an overview of **`r colorise("janitor", "blue")`**, an R package that helps beginner/intermediate R users to **format columns, clean dirty data, and explore data sets** in order to easily and quickly prepare data for analysis! `r colorise("janitor", "blue")` follows tidyverse principles and works well with piping.

```{r, message=F, comment=F}
library(tidyverse)
library(janitor)
library(readxl)
```

## Importing data

We can **import our excel file** using the `read_excel()` function.

```{r, message=F, comment=F}
dirty_data <- read_excel("sample_data.xlsx")
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Cleaning column names

We use `clean_names()` to **manage column names**. The function takes all column names and converts to lowercase letters, adds "_" as a separator, gets rid of special characters and unnecessary spaces, assigns numbers to duplicated names, and converts “%” to “percent”.

```{r, message=F, comment=F}
dirty_data

student_data <- dirty_data %>% clean_names()
student_data

data.frame(dirty_data = colnames(dirty_data), student_data = colnames(student_data))
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Removing redundant rows/columns

We use `remove_empty()` to **get rid of empty rows and columns**, and use `remove_constant()` to **get rid of columns with constant values**.

```{r, message=F, comment=F}
student_data

student_data <- student_data %>% remove_empty(c("rows", "cols"))
student_data

student_data <- student_data %>% remove_constant()
student_data
```

## Rounding

This packages offer **multiple rounding functions** which we can utilize for our specific needs. We can use `round_half_up()` for rounding up or `round_to_fraction()` to enforce the desired fractional distribution by rounding the values to the nearest value by specifying a denominator.

```{r, message=F, comment=F}
round_half_up(student_data$gpa)

round_half_up(student_data$gpa, digits = 2)

round_to_fraction(student_data$gpa, denominator = 4)
```

## Fixing date formats

We can also easily **fix dates** that do not appear in proper form when imported from external files using `convert_to_date()`.

```{r, message=F, comment=F}
student_data

student_data$latest_vaccination <- convert_to_date(student_data$latest_vaccination)
student_data
```
## Getting dupes

We can use the `get_dupes()`function to **detect any duplicates**, have our data frame listed on the specified column, and get dupe counts.

```{r, message=F, comment=F}
student_data %>% get_dupes()

student_data %>% get_dupes(first_name)

student_data %>% get_dupes(vaccine)
```

## Exploring data through frequency

The function `tabyl()` **returns neat and quick frequency tables**. Later, we can utilize a list of `adorn_xxx()` functions to format displayed results.

```{r, message=F, comment=F}

student_data %>%
  tabyl(declared_major) %>%
  adorn_totals("row") %>%
  adorn_pct_formatting()
  
student_data %>%
  tabyl(age, class_of) %>%
  adorn_totals("row") %>%
  adorn_percentages() %>%
  adorn_pct_formatting() %>%
  adorn_ns() %>%
  adorn_title()
```

## A few more useful tricks

What if we have couple of rows we want to discard? We can use `row_to_names()` function to **pick any row as our column names**.

```{r, message=F, comment=F}
library_data <- data.frame(
  Library_Stats_1 = c("Check outs from the library", "Number of books", "8", "12", "4"),
  Library_Stats_2 = c("Number of completed books", "Read", "5", "9", "2"))

library_data

row_trick <- library_data %>% row_to_names(row_number = 2)

row_trick
```

We can also **observe the changes we have made** to the columns by using the `compare_df_cols()` function.

```{r, message=F, comment=F}
compare_df_cols(library_data, row_trick)
```
## Sources

This tutorial draws from the following materials:
[_janitor package_](https://cran.r-project.org/web/packages/janitor/janitor.pdf), [_overview of janitor functions_](https://cran.r-project.org/web/packages/janitor/vignettes/janitor.html#tabyl---a-better-version-of-table), [_catalog of janitor functions_](https://garthtarr.github.io/meatR/janitor.html#catalog_of_janitor_functions), [_R package reviews | janitor | clean your data!_](https://www.youtube.com/watch?v=poUveR6aDS8)

